---
title: "Compound Poisson-Normal Regression"
author: "Laszlo Pecze"
date: "`r Sys.Date()`"
bibliography: references.bib
nocite: |
  @Nascimento2019, @hu2024, @raqab2021
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Compound Poisson-Normal Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```




## Introduction

In many applied settings—including insurance claims, biological systems, and environmental monitoring—the observed data arise from a random number of additive events, where each event contributes a continuous value. A natural model for such data is the **Compound Poisson-Normal (CPN)** distribution.

The CPN model assumes that the number of events follows a Poisson distribution, and that each event contributes an independent and identically distributed (i.i.d.) **normally distributed** amount.

Formally, let:

- \( N \sim \text{Poisson}(\lambda) \) denote the number of events,
- \( X_1, X_2, \ldots, X_N \overset{\text{i.i.d.}}{\sim} \mathcal{N}(\mu, \sigma^2) \), independent of \( N \),

Then the total observed outcome is modeled as a **compound sum**:

\[
Y = \sum_{i=1}^{N} X_i
\]

The resulting distribution of \( Y \) is known as the **Compound Poisson-Normal distribution**.


The first two moments of the CPN distribution are:

\[
\mathbb{E}[Y] = \lambda \mu,\quad \text{Var}(Y) = \lambda (\mu^2 + \sigma^2)
\]

These follow from the properties of the compound distribution, combining Poisson and Gaussian contributions.
The (approximate) probability density function (pdf) of \( Y \) is:

\[
f_Y(y; \lambda, \mu, \sigma) =
\begin{cases}
e^{-\lambda}, & \text{if } y = 0 \\\\
\sum_{k=1}^{K_{\text{max}}} \frac{e^{-\lambda} \lambda^k}{k!} \cdot \frac{1}{\sqrt{2\pi k \sigma^2}} \exp\left( -\frac{(y - k\mu)^2}{2k\sigma^2} \right), & \text{if } y \neq 0
\end{cases}
\]

**Where:**

- \( y \): the observed total from the compound process  
- \( \lambda \): expected number of Poisson events  
- \( \mu \): mean of each normal component  
- \( \sigma \): standard deviation of each normal component  
- \( k \): number of events (from 1 to \( K_{\text{max}} \))  
- \( K_{\text{max}} \): maximum number of events used in the approximation  

When \( k = 0 \), the sum is defined as a point mass at zero, and its probability is:

\[
P(Y = 0) = P(N = 0) = e^{-\lambda}
\]

In theory, the compound distribution sums over an infinite range of possible Poisson event counts. In practice, this sum must be **truncated** at a finite maximum value \( K_{\max} \), chosen so that the remaining tail probability is negligible.

## Getting Started

To use the function, load the package:

```{r library}
library(CPN)
```

## Simulating Data

We’ll simulate a dataset that mimics typical CPN behavior. This includes both categorical and continuous predictors, and outcomes generated via a Poisson count of normally distributed values.

```{r simu0, message=FALSE, warning=FALSE, include=FALSE}
simulate_cpn_data <- function(n = 100,
                              beta = c(0.5, -0.3, 0.7),
                              mu = 1,
                              sigma = 2) {
  set.seed(123)  # For reproducibility

  # Simulate predictors
  x1 <- factor(sample(c("A", "B"), size = n, replace = TRUE))  # Categorical
  x2 <- rnorm(n)  # Continuous

  # Create model matrix (includes intercept and dummy variable for x1)
  X <- model.matrix(~ x1 + x2) # nolint

  # Compute linear predictor and Poisson rates
  eta <- X %*% beta
  lambda <- exp(eta)

  # Simulate response variable
  y <- numeric(n)
  for (i in 1:n) {
    k <- stats::rpois(1, lambda[i])
    if (k > 0) {
      y[i] <- sum(stats::rnorm(k, mean = mu, sd = sigma))
    } else {
      y[i] <- 0
    }
  }

  # Return as a data.frame
  data.frame(y = y, x1 = x1, x2 = x2)
}
```

```{r simu, message=FALSE, warning=FALSE}
set.seed(123)
data <- simulate_cpn_data()
head(data)
```


## Plot the simulated data

Visualize the response y against predictor x2, colored by group x1.
```{r simuplot, message=FALSE, warning=FALSE}
# Scatter plot of y vs x1
stripchart(
  y ~ x1, data = data,
  vertical = TRUE, method = "jitter",
  pch = 19, cex = 0.6,
  col = c("blue",  "red"),
  xlab = "x1", ylab = "y",
  main = "Scatter Plot of y by x1"
)

```

##  Fit the CPN Model
Fit a Compound Poisson-Normal regression model to the data using a standard formula interface. The response variable y is modeled as a function of predictors x1 (categorical) and x2 (continuous).
```{r}
fit <- cpn(y ~ x1 + x2, data = data)
```

##  Summary of Results
Get a summary of the fitted model, including coefficient estimates, standard errors, z-values, and p-values. The summary also includes estimates of the mu and sigma parameters (mean and SD of the normal component) and model fit statistics like AIC.
```{r}
summary(fit)
```

##  Extracting Model Components
You can directly access key components of the fitted model object:
```{r}
fit$coefficients
fit$mu
fit$sigma
fit$fitted_values[1:10]  # preview only
```

##  Coefficients and Interpretation
Use coef() to extract model coefficients. Setting full = FALSE shows only the linear predictor coefficients (excluding auxiliary parameters like mu and sigma).

```{r coef}
coef(fit)
coef(fit, full = FALSE)  # Includes only linear predictors
```

##  Diagnostics and Residuals
Basic diagnostic plots and residuals help check model fit and detect issues like non-linearity or outliers.
```{r diagnostics}
plot(fit)
residuals(fit)[1:10]
```

##  Likelihood and Information Criteria
These metrics are useful for comparing model fit. `logLik()` gives the log-likelihood, `AIC()` the Akaike Information Criterion (lower is better), and `vcov()` returns the variance-covariance matrix of the parameters.

```{r fitstats}
logLik(fit)
AIC(fit)
vcov(fit)
```

## Type I Analysis of Deviance

The function `anova(fit)` provides a sequential (Type I) analysis of deviance.  
This examines the incremental contribution of each predictor to the model's fit, based on the order in which they appear in the formula.

```{r anova}
anova(fit)
```

## Model Update and Comparison
You can update a model (e.g., remove predictors) and compare nested models using the likelihood ratio test:
```{r update-anova}
fit2 <- update(fit, . ~ . - x2)

anova(fit, fit2)  # Likelihood ratio test
```


## Prediction

Once you have fitted a Compound Poisson-Normal (CPN) regression model using the `cpn()` function, you can generate predictions on both the **original data** and **new observations** using the `predict()` method for `cpn` objects.

This method supports three prediction types:

- **"link"**: Returns the linear predictor \(\eta = X\beta\).
- **"rate"**: Returns the rate component \(\exp(\eta)\), representing the expected number of latent events.
- **"response"**: Returns the expected response \(\mathbb{E}[Y] = \mu \cdot \exp(\eta)\), combining the frequency and magnitude components of the compound distribution.

Additionally, confidence intervals can be requested using the `interval = "confidence"` argument, which computes approximate **normal-theory confidence intervals** using the **delta method**.

### Predicting on the Original Data

By default, `predict()` uses the original dataset used for model fitting. You can obtain point estimates or confidence intervals for the fitted values.

```{r predict-original}
# Point predictions on the response scale
predict(fit, type = "response")[1:5]

# Confidence intervals for response predictions
predict(fit, type = "response", interval = "confidence")[1:5, ]
```

### Predicting on New Data

To make predictions on new data, provide a `data.frame` to the `newdata` argument. The function will internally align factor levels and construct the appropriate model matrix.

```{r predict-new}
# Define new observations
new_df <- data.frame(
  x1 = c("A", "A", "B", "B"),
  x2 = c(-0.5, -0.2, -0.3, -0.3)
)

# Predictions on the link scale (linear predictor η)
predict(fit, newdata = new_df, type = "link", interval = "confidence")

# Predictions on the rate scale (exp(η))
predict(fit, newdata = new_df, type = "rate", interval = "confidence")

# Predictions on the response scale (μ × exp(η))
predict(fit, newdata = new_df, type = "response", interval = "confidence")
```

### Confidence Intervals

When `interval = "confidence"` is used:

- On the **link** and **rate** scales, confidence intervals are based solely on the uncertainty in the regression coefficients (\(\beta\)).
- On the **response** scale, the intervals additionally incorporate uncertainty in the mean magnitude parameter (\(\mu\)) using the **delta method**:

\[
\text{SE}[\mu \cdot \exp(\eta)] \approx \sqrt{(\mu \cdot \exp(\eta))^2 \cdot \text{Var}[\eta] + (\exp(\eta))^2 \cdot \text{Var}[\mu]}
\]

This provides a more accurate reflection of total uncertainty when predicting expected values of the compound outcome.

```

## Conclusion

The `CPN` package enables effective modeling of semicontinuous data using the
Compound Poisson-Normal model. It supports flexible model specification, S3
methods for diagnostics and inference, and prediction on new data. For more
advanced use, refer to the function documentation or the package reference
manual.

##  References
